Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-06 manuscript/manuscript.tex
algo
corrections

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}

@@

\textbf{\textit{Put lit review of Bayesian multinomial logistic regression here}}

@@

\[   P(Z_i = k) = \left \{
\begin{array}{ll}
      \pi_{ik} = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{1 + \sum_{k' = 1}^{K-1} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}} & k < K \\
      \\
      \pi_{ik} = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{1 + \sum_{k' = 1}^{K-1} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}} & k = K \\
\end{array} 
\right. \]

@@

\subsection{Conditional MVSN Imputation}

@@

This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{k - q \times 1}}]^T$. We have

@@

@@

@@

    \item $\mathbf{Y}$: A $N \times J$ matrix containing all multivariate skew-normal  outcomes such that $y_{ij}$ is the $j^{th}$ outcome observed for subject $i$, where $i = 1,...,n$ and $j = 1,...J$.

    \item $\mathbf{X}$: A $n \times p$ matrix containing all multivariate skew-normal regression covariates such that $x_{ij}$ is the $j^{th}$ covariate value for subject $i$.

    \item $\mathbf{B}$: A $m \times p$ matrix containing all multivariate skew-normal regression coefficients such that $\mathbf{B} = \left [ \boldsymbol\beta_1,...,\boldsymbol\beta_p \right ]$, where $\beta_{ij}$ is interpreted as the effect of covariate $i$ on outcome $j$ for $i = 1,...,m$ and $j = 1,...,p$.

    \item $\mathbf{E}$: A $n \times p$ matrix of error terms in the multivariate skew-normal regression model component. $\mathbf{E}$ is made up of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{ip})$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_p(0, \boldsymbol\Sigma)$ for $i = 1,...,n$.

    \item $\boldsymbol\Sigma$: A $p \times p$ covariance matrix that defines the correlation between the $p$ multivariate normal outcomes.

    \item $\boldsymbol\Omega$: A $p \times p$ covariance scale matrix that defines the correlation between the $p$ multivariate skew-normal outcomes.

    \item $\boldsymbol\psi$: A $p \times 1$ vector containing the skewness parameter for each outcome.

    \item $\boldsymbol\alpha$: A $p \times 1$ vector containing the skewness parameter for each outcome.

@@

    \item $\mathbf{X}^*$: A $n \times (m + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$

    \item $\mathbf{B}^*$: A $(m+1) \times p$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$.

@@

The multivariate skew-normal regression model can be written as follows in matrix form. 
$$\mathbf{Y} = \mathbf{X}\mathbf{B} + \mathbf{t} \boldsymbol\psi^T + \mathbf{E} = \mathbf{X}^* \mathbf{B}^* + \mathbf{E}$$
The matrix $\mathbf{Y}$ is of dimension $n \times p$. For convenience, we define $\mathbf{X}^*$ as a $n \times (m + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$, and $\mathbf{B}^*$ as a $(m+1) \times p$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$. We assume that $t_i \stackrel{iid}{\sim}T_{[0,\infty})(0,1)$ and that $\mathbf{E}$ is made of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{ip})$ for $i = 1,...,n$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_p(0, \boldsymbol\Sigma)$.

@@

We choose conjugate priors for $\mathbf{B}^*$ and $\boldsymbol\Sigma$ as follows. 
$$\boldsymbol\Sigma \sim \text{inverse-Wishart}(\mathbf{V}_0,\nu_0)$$

$$\mathbf{B}^*|\boldsymbol\Sigma \sim MatNorm_{(m+1), p}(\mathbf{B}_0^*,\mathbf{L}_0^{-1},\boldsymbol\Sigma)$$

We now dervive the joint posterior distribution of the parameters $\mathbf{B}^*$ and $\boldsymbol\Sigma$.
\begin{align*} 
p(\mathbf{B}^*,\boldsymbol\Sigma|\mathbf{X}^*,\mathbf{Y}) & \propto  p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma)p(\mathbf{B}^*|\boldsymbol\Sigma)p(\boldsymbol\Sigma)\\
&  \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{-(m+1)/2} \exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{B}^* - \mathbf{B}^*_0)^T \mathbf{L}_0(\mathbf{B}^* - \mathbf{B}^*_0)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{(\nu_0 + p + 1)/2} \exp \left \{ -\frac{1}{2} \tr (\mathbf{V}_0 \boldsymbol\Sigma^{-1}) \right \}
\end{align*}

@@

For a given observation vector $\mathbf{y} \sim N_p(\boldsymbol\mu, \boldsymbol\Sigma)$, we allow for missingness in at most $p - 1$ of the multivariate outcomes through the use of a conditional imputation step embedded within our Gibbs sampler. Suppose $\mathbf{y}$ contains $q$ missing observations and can be partitioned into two vectors $\mathbf{y_{1}}$ and $\mathbf{y_{2}}$ such that $\mathbf{y_{1}}$ is a $q \times 1$ vector of missing observations and $\mathbf{y_{2}}$ is a $(p-q) \times 1$ vector of complete observations. Similarly, partition $\boldsymbol\mu$ and $\boldsymbol\Sigma$ as follows.
$$\boldsymbol\mu = \begin{bmatrix} \boldsymbol\mu_1 \\ \boldsymbol\mu_2 \end{bmatrix} \ \ \ \ \ \boldsymbol\Sigma = \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}$$

@@

>>>>

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dsfont}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

@@

\textbf{\textit{Put lit review of Bayesian PG multinomial logistic regression here}}

@@

\[   P(Z_i = k|\mathbf{w}_i,\boldsymbol\delta_k) = \left \{
\begin{array}{ll}
      \pi_{ik} = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{1 + \sum_{k' = 1}^{K-1} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}} & k < K \\
      \\
      \pi_{ik} = \frac{1}{1 + \sum_{k' = 1}^{K-1} e^{\mathbf{w}_i^T \boldsymbol\delta_{k'}}} & k = K \\
\end{array} 
\right. \]

@@

\subsection{Conditional MVN Imputation}

@@

This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{J - q \times 1}}]^T$. We have

@@

\textit{\textbf{Discuss non-ignorable missingness here}}

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n \mathds{1}_{Z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(Z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{Z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ Z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{Z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

@@

    \item $\mathbf{Y}$: A $n \times J$ matrix containing all multivariate skew-normal  outcomes such that $y_{ij}$ is the $j^{th}$ outcome observed for subject $i$, where $i = 1,...,n$ and $j = 1,...J$.

    \item $\mathbf{X}$: A $n \times P$ matrix containing all multivariate skew-normal regression covariates such that $x_{ip}$ is the $p^{th}$ covariate value for subject $i$, where $i = 1,...,n$ and $p = 1,...P$.

    \item $\mathbf{B}$: A $P \times J$ matrix containing all multivariate skew-normal regression coefficients such that $\mathbf{B} = \left [ \boldsymbol\beta_1,...,\boldsymbol\beta_J \right ]$, where $\beta_{pj}$ is interpreted as the effect of covariate $p$ on outcome $j$ for $p = 1,...,P$ and $j = 1,...,J$.

    \item $\mathbf{E}$: A $n \times J$ matrix of error terms in the multivariate skew-normal regression model component. $\mathbf{E}$ is made up of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{iJ})$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_J(0, \boldsymbol\Sigma)$ for $i = 1,...,n$.

    \item $\boldsymbol\Sigma$: A $J \times J$ covariance matrix that defines the correlation between the $p$ multivariate normal outcomes.

    \item $\boldsymbol\Omega$: A $J \times J$ covariance scale matrix that defines the correlation between the $p$ multivariate skew-normal outcomes.

    \item $\boldsymbol\psi$: A $J \times 1$ vector containing the skewness parameter for each outcome.

    \item $\boldsymbol\alpha$: A $J \times 1$ vector containing the skewness parameter for each outcome.

@@

    \item $\mathbf{X}^*$: A $n \times (P + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$

    \item $\mathbf{B}^*$: A $(P+1) \times J$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$.

@@

The multivariate skew-normal regression model can be written as follows in matrix form. 
$$\mathbf{Y} = \mathbf{X}\mathbf{B} + \mathbf{t} \boldsymbol\psi^T + \mathbf{E} = \mathbf{X}^* \mathbf{B}^* + \mathbf{E}$$
The matrix $\mathbf{Y}$ is of dimension $n \times J$. For convenience, we define $\mathbf{X}^*$ as a $n \times (P + 1)$ matrix constructed by column binding $\mathbf{t}$ to $\mathbf{X}$, and $\mathbf{B}^*$ as a $(P+1) \times J$ matrix constructed by row binding $\boldsymbol\psi^T$ to $\mathbf{B}$. We assume that $t_i \stackrel{iid}{\sim}T_{[0,\infty})(0,1)$ and that $\mathbf{E}$ is made of row vectors $\boldsymbol\epsilon_i = (\epsilon_{i1},...,\epsilon_{iJ})$ for $i = 1,...,n$, where $ \boldsymbol\epsilon_i \stackrel{iid}{\sim} N_J(0, \boldsymbol\Sigma)$.

@@

We choose conjugate priors for $\mathbf{B}^*$ and $\boldsymbol\Sigma$ as follows. 
$$\boldsymbol\Sigma \sim \text{inverse-Wishart}(\mathbf{V}_0,\nu_0)$$
$$\mathbf{B}^*|\boldsymbol\Sigma \sim MatNorm_{(m+1), p}(\mathbf{B}_0^*,\mathbf{L}_0^{-1},\boldsymbol\Sigma)$$

We now dervive the joint posterior distribution of the parameters $\mathbf{B}^*$ and $\boldsymbol\Sigma$.
\begin{align*} 
p(\mathbf{B}^*,\boldsymbol\Sigma|\mathbf{X}^*,\mathbf{Y}) & \propto  p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma)p(\mathbf{B}^*|\boldsymbol\Sigma)p(\boldsymbol\Sigma)\\
&  \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{-(P+1)/2} \exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{B}^* - \mathbf{B}^*_0)^T \mathbf{L}_0(\mathbf{B}^* - \mathbf{B}^*_0)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{(\nu_0 + J + 1)/2} \exp \left \{ -\frac{1}{2} \tr (\mathbf{V}_0 \boldsymbol\Sigma^{-1}) \right \}
\end{align*}

@@

For a given observation vector $\mathbf{y} \sim N_J(\boldsymbol\mu, \boldsymbol\Sigma)$, we allow for missingness in at most $J - 1$ of the multivariate outcomes through the use of a conditional imputation step embedded within our Gibbs sampler. Suppose $\mathbf{y}$ contains $q$ missing observations and can be partitioned into two vectors $\mathbf{y_{1}}$ and $\mathbf{y_{2}}$ such that $\mathbf{y_{1}}$ is a $q \times 1$ vector of missing observations and $\mathbf{y_{2}}$ is a $(J-q) \times 1$ vector of complete observations. Similarly, partition $\boldsymbol\mu$ and $\boldsymbol\Sigma$ as follows.
$$\boldsymbol\mu = \begin{bmatrix} \boldsymbol\mu_1 \\ \boldsymbol\mu_2 \end{bmatrix} \ \ \ \ \ \boldsymbol\Sigma = \begin{bmatrix} \boldsymbol\Sigma_{11} & \boldsymbol\Sigma_{12} \\ \boldsymbol\Sigma_{21} & \boldsymbol\Sigma_{22} \end{bmatrix}$$

@@

\bibitem{ } Gelman A, Hwang J, Vehtari A. Understanding predictive information criteria for Bayesian models. \textit{Statistics and Computing}. 2014 Nov 1;24(6):997-1016.
